Storm surge modelling in the Strait of Georgia

### 2014-07-24 RoughWork/manuscript.tex
Worked on modelling overview in intro

\begin{abstract}
The Strait of Georgia is a large, semi-enclosed body of water between Vancouver Island and the mainland of British Columbia connected to the Pacific Ocean via the Strait of Juan de Fuca at the south and Johnstone Strait at the north. During the winter months, coastal communities in the Strait of Georgia are particularly susceptible to flooding caused by storm surges, a natural hazard that can occur when a strong storm coincides with high tide. This investigation produces storm surge hindcasts using a three-dimensional numerical ocean model for the Strait of Georgia and the surrounding bodies of water (Strait of Juan de Fuca, Puget Sound, and Johnstone Strait) collectively known as the Salish Sea. The numerical model employs the Nucleus for European Modelling of the Ocean (NEMO) architecture in a regional configuration. The model is evaluated through comparisons of tidal elevation harmonics, tidal current ellipses, and storm surge water levels with observations. Important forcing factors contributing to storm surges have been assessed, finding that the surge entering the domain from the Pacific Ocean contributes most significantly. In addition, local wind patterns cause a slight increase in surge amplitude on the mainland side of the Strait of Georgia when compared to the Vancouver Island coastal areas. Surge amplitudes are found to be greater within the Strait of Georgia when compared to amplitudes in the Strait of Juan de Fuca. 
%needs a little more work.
\end{abstract}

@@

%some general intro paragraph here
%Need something about SoG size, geography, islands, sills, rivers....

%my attempt at physical oceanopgraphy in SoG- Susan feel free to edit.
The Strait of Georgia is a strongly stratified, semi-enclosed body of water located between Vancouver Island and the mainland of British Columbia. It is part of a larger system of waterways collectively known as the Salish Sea and is connected to the Pacific Ocean via the Strait of Juan de Fuca and Johnstone Strait. Several features related to the physical oceanopgraphy of this region render it a challenge to model numerically. The outflow from the Fraser River in the late spring is the primary source fresh water within the Strait of Georgia and leads to large vertical density gradients in the summer months. Additionally, strong tidal mixing through the San Juan and Gulf Islands produces horizontal density gradients that separate the saline waters of the Strait of Juan de Fuca and the fresher water of the Strait of Georgia. It is important to accurately represent the vertical mixing in this region as it sets the rate of export of fresh water through estuarine circulation. The strong and fast tidal currents through the narrow channels of the north, such as Discovery Passage and Johnstone Strait, pose a particular challenge for numerical models. Yet, several modelling efforts for this region exist, ranging in choices of grid structure, mixing parametrizations, and domain size and extent.

%overview oF SoG/Salish Sea models
Early work by \citet{crean1988numerical, crean1988mathematical, stronach1993update} resulted in a series of models covering the Strait of Juan de Fuca, the Strait of Georgia and Puget Sound using a finite difference discretization of the equations of motion. A barotropic version of their model was used to reproduce storm surge water level elevations with reasonable accuracy, however, this barotropic version did not include the effects of fresh water entering the system from rivers. Later, a baroclinc version of their model examined residual currents and more. %elaborate!
 The baroclinic model of \citet{masson2004modelling} examined seasonal variability of deep water properties in the Strait of Georgia noting a strong sensitivity to vertical mixing over the sills. This model employed a sigma vertical coordinate system leading to errors in diapycnal mixing due to steep topographics slopes. %fix this up still.
%tide models (Foreman), baroclinic (Masson, PNWL, Sutherland and MacCready)

@@

%overview of paper: model configuration, model evaluation, storm surge hindcasts.
This paper is organized as follows: an overivew of the model configuration and domain including a description of open and surface boundary conditions is provided in section \ref{sec:config}. This section also describes the ocean model we have employed as well as choices in grid structure and mixing parameterizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's accuracy in producing storm surge hindcasts is also assesed in section \ref{sec:storm}. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{Figures/bathy.pdf}
\caption{Model domain (highlighted area) including bathymetry, rivers (blue circles), and storm surge locations of interest  (red stars).}\label{fig:domain}
\end{figure}
%image needs some work: labels for tide comparisons?

@@

Finally, an assessment of the importance of including the tide-surge interaction is presented next. A simulation of the surge propagation without including any tidal forcing is compared with the modelled residual in Figure \ref{fig:tidesurge}.  Recall, the modelled residual is defined as the difference between the sea surface height in a simulation with all forcing and a simulaltion with tides amd rivers only. A comparison between the modelled residual and a surge-only run can determine the significance of the tide-surge interaction. The most significant differences occur at Point Atkinson where the maximum modelled residual is 7 cm lower than the maximum of the surge-only case.  In contrast, the tide-surge interaction at Victoria is hardly noticeable. Indeed, a spatial map of the modelled residual and surge-only differences (not shown here) indicate that the tide-surge interaction is more pronounced after the San Juan and Gulf Islands. The passages between these islands are known for vigorous tidal mixing, which may have an effect on the surge propagation. However, the effect is a relatively small underestimate of the surge elevation and no changes to the timing of the surge. So, in operational settings, performing a surge-only simulation and adding the predicted tides a posteriori could be more computationally efficient with only a small decrease in the accuracy of the prediction.

@@

Point Atkinson observed a maximum surge amplitude of 84 cm, which is approximately 20 cm higer than the model result, however the timing of the surge agrees almost exactly. The best agreement occurs at Victoria where the observed surge was 62 cm and the model returned 54 cm. At all of the locations, the model predicted the maximum surge to occur within 3.5 hours of the observed surge. Typically, the modelled surge was too early, in contrast to the later predictions in the Feb, 4 2006 case. Errors in the modelled water level at the time of maximum surge were of the same order as the errors in the surge amplitude.

>>>>

\begin{abstract}
The Strait of Georgia is a large, semi-enclosed body of water between Vancouver Island and the mainland of British Columbia connected to the Pacific Ocean via the Strait of Juan de Fuca at the south and Johnstone Strait at the north. During the winter months, coastal communities in the Strait of Georgia are at risk to flooding caused by storm surges, a natural hazard that can occur when a strong storm coincides with high tide. This investigation produces storm surge hindcasts using a three-dimensional numerical ocean model for the Strait of Georgia and the surrounding bodies of water (Strait of Juan de Fuca, Puget Sound, and Johnstone Strait) collectively known as the Salish Sea. The numerical model employs the Nucleus for European Modelling of the Ocean (NEMO) architecture in a regional configuration. The model is evaluated through comparisons of tidal elevation harmonics, tidal current ellipses, and storm surge water levels with observations. Important forcing factors contributing to storm surges have been assessed, finding that the surge entering the domain from the Pacific Ocean contributes most significantly. In addition, local wind patterns cause a slight increase in surge amplitude on the mainland side of the Strait of Georgia when compared to the Vancouver Island coastal areas. Surge amplitudes are found to be greater within the Strait of Georgia when compared to amplitudes in the Strait of Juan de Fuca. 
%needs a little more work.
\end{abstract}

@@

%some general intro paragraph here. Need something about SoG size, geography, islands, sills, rivers, tides....
The Strait of Georgia is a strongly stratified, semi-enclosed body of water located between Vancouver Island and the mainland of British Columbia. It is part of a larger system of waterways collectively known as the Salish Sea and is connected to the Pacific Ocean via the Strait of Juan de Fuca and Johnstone Strait.

%my attempt at physical oceanopgraphy in SoG- Susan feel free to edit.
Several features related to the physical oceanopgraphy of this region render it a challenge to model numerically. The outflow from the Fraser River in the late spring is the primary source fresh water within the Strait of Georgia and leads to large vertical density gradients in the summer months. Additionally, strong tidal mixing through the San Juan and Gulf Islands produces horizontal density gradients that separate the saline waters of the Strait of Juan de Fuca and the fresher water of the Strait of Georgia. It is important to accurately represent the vertical mixing in this region as it sets the rate of export of fresh water through estuarine circulation. The strong and fast tidal currents through the narrow channels of the north, such as Discovery Passage and Johnstone Strait, pose a particular challenge for numerical models. Yet, several modelling efforts for this region exist, ranging in choices of grid structure, mixing parametrizations, and domain size and extent.

%overview oF SoG/Salish Sea models
Early work by \citet{crean1988mathematical,crean1988numerical} and \citet{stronach1993update} resulted in a series of models covering the Strait of Juan de Fuca, the Strait of Georgia and Puget Sound using a finite difference discretization of the equations of motion. A barotropic version of their model was used to reproduce storm surge water level elevations with reasonable accuracy, however, this barotropic version did not include the effects of fresh water entering the system from rivers. Later, a baroclinc version of their model examined residual circulation and fresh water flushing rates \citep{stronach1993update, marinone1996model}. This early work pointed to an important connection between the model's bottom friction parameterization and the location and amplitude of the M$_2$ amphidrome. They found that a local increase in the bottom friction coefficient, especially in the area of Haro Strait, was necessary in order to adequately reproduce the observed tidal amplitudes and phase lags in the Strait of Georgia \citep{stronach1993update}. Other studies over a similar domain, such as the barotropic finite element model of \citet{foreman1995tidal}, have avoided local increases in bottom friction. \citet{foreman2004m} used data assimilation techinques to study M$_2$ tidal dissipation without local changes to the bottom frction in a two-dimensional finite element model.

More recently, modelling studies have focused on baroclinic features. \citet{masson2004modelling} used the Princeton Ocean Model (POM) in a regional mode to examine the seasonal variability of deep water properties in the Strait of Georgia noting a strong sensitivity to vertical mixing over the sills. This model employed a sigma or terrain-following vertical coordinate system leading to spurious diapycnal mixing since, numerically, the lateral mixing is computed on sigma levels which may intersect isopycnals in the presence of steep topography. \citet{yang2010multi} developed a model of southern Strait of Georgia and Puget Sound using volume techniques on an unstructured grid. \citet{sutherland2011model} employed the Regional Ocean Modeling System (ROMS) to study estuarine circulation in the Salish Sea with a model domain that extended outside the Strait of Juan de Fuca into the Pacific Ocean and up to the northern Strait of Georgia. The passageways through the Discovery Islands north of the Strait of Georgia have been modelled baroclinically by \citet{foreman2012circulation} with the Finite Volume Coastal Ocean Model (FVCOM) pointing to difficulties in reproducing baroclinic currents on their grid.  These studies have led to an improvement in understanding circulation properties in the Salish Sea and have highlighted important features that should be considered in future modelling work.

@@

%overview of paper: model configuration, model evaluation, storm surge hindcasts.
This paper is organized as follows: an overview of the model configuration and domain including a description of the ocean model we have employed  is provided in section \ref{sec:config}. This section also describes open and surface boundary conditions as well as choices in grid structure and mixing parameterizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's accuracy in producing storm surge hindcasts is also assesed in section \ref{sec:storm}. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{Figures/bathy.pdf}
\caption{Model domain (highlighted area) including bathymetry, rivers (blue circles), and storm surge locations of interest  (red stars).}\label{fig:domain}
\end{figure}
%image may need more work

@@

Finally, an assessment of the importance of including the tide-surge interaction is presented next. A simulation of the surge propagation without including any tidal forcing is compared with the modelled residual in Figure \ref{fig:tidesurge}.  Recall, the modelled residual is defined as the difference between the sea surface height in a simulation with all forcing and a simulaltion with tides amd rivers only. A comparison between the modelled residual and a surge-only run can determine the significance of the tide-surge interaction. The most significant differences occur at Point Atkinson where the maximum modelled residual is 7 cm lower than the maximum of the surge-only case.  In contrast, the tide-surge interaction at Victoria is hardly noticeable. Indeed, a spatial map of the modelled residual and surge-only difference (not shown here) indicates that the tide-surge interaction is more pronounced after the San Juan and Gulf Islands. The passages between these islands are known for vigorous tidal mixing, which may have an effect on the surge propagation. However, the effect is a relatively small underestimate of the surge elevation and no changes to the timing of the surge. So, in operational settings, performing a surge-only simulation and adding the predicted tides a posteriori could be more computationally efficient with only a small decrease in the accuracy of the prediction.

@@

Point Atkinson observed a maximum surge amplitude of 84 cm, which is approximately 20 cm higher than the model result, however the timing of the surge agrees almost exactly. The best agreement occurs at Victoria where the observed surge was 62 cm and the model returned 54 cm. At all of the locations, the model predicted the maximum surge to occur within 3.5 hours of the observed surge. Typically, the modelled surge was too early, in contrast to the later predictions in the Feb 4, 2006 case. Errors in the modelled water level at the time of maximum surge were of the same order as the errors in the surge amplitude.
